---
title: "Pharmacogenomics Lab"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Project Goals
1. How does Gene ABCB1 associate with different diseases?
2. How strong is the evidence for the association between Variant X and different disease?  

##Resources  
- [Warfarin data](https://www.pharmgkb.org/downloads)  

## Load the libraries
```{r}
library(tidyverse)
library(janitor)
```

## Dataa
```{r}
relations <- read_tsv("data/relationships.tsv") %>% clean_names()
```

```{r}
names(relations)
glimpse(relations)
```

#Percentage of association between regular genes and diseases
```{r}
pa_gene_disease <- relations %>% 
  filter(entity1_type == "Gene" & entity2_type == "Disease") %>% 
  group_by(association) %>%
  summarise(count = n(), .groups = 'drop') %>%
  mutate(percentage = (count / sum(count)) * 100) %>% 
  arrange(-count)
pa_gene_disease
```


#Graph of percentage of association between variant genes and diseases
```{r}
ggplot(pa_gene_disease, aes(x = "", y = percentage, fill = association)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  labs(title = "Percentage of Association Types between Genes and Diseases",
       fill= "Type of Association") +
  scale_fill_brewer(palette = "Paired")+
  theme_void()
```

1. Find the most common gene as entity 1.
```{r}
relations %>% 
  filter(entity1_type == "Gene") %>% 
  count(entity1_name, sort = T)

# The most common gene appears to be ABCB1! (gene with the most data comparisons)
```

2. Filter the data set to have only the data we want.
```{r}
filtered_relations <- relations %>% 
  select(-pk, -pd) %>% 
  filter(entity1_name == "ABCB1" & entity2_type == "Disease" & association == "associated")
```

3. Count the number of diseases
```{r}
disease_count <- filtered_relations %>% 
  count(entity2_name, sort = T)
```

4. Visualize the significance. #should we get rid of this one?
```{r}
ggplot(disease_count, aes(x = reorder(entity2_name, n), y = n)) +
  geom_bar(stat = "identity", fill = "hotpink") +
  coord_flip() +
  labs(title = "Diseases Associated with ABCB1",
       x = "Disease",
       y = "Number of Associations") +
  theme_minimal()
```

5. Categorize the diseases
```{r}
disease_count <- disease_count %>%
  mutate(Category = case_when(
    grepl("cancer|carcinoma|lymphoma|tumor", entity2_name, ignore.case = TRUE) ~ "Cancer",
    grepl("diabetes", entity2_name, ignore.case = TRUE) ~ "Metabolic",
    grepl("hypertension|cardiac|coronary|arrhythmia", entity2_name, ignore.case = TRUE) ~ "Cardiovascular",
    grepl("autism|attention deficit|nervous system", entity2_name, ignore.case = TRUE) ~ "Neurological",
    TRUE ~ "Other"
  ))

# Count number of diseases per category
category_counts <- disease_count %>%
  count(Category, sort = TRUE)

# Plot disease categories
ggplot(category_counts, aes(x = reorder(Category, n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Disease Categories Associated with ABCB1",
       x = "Disease Category",
       y = "Number of Associations") +
  theme_minimal()

# ABCB1 is associated with a wide variety of diseases that don’t fit into common categories like cancer, cardiovascular, neurological, or metabolic diseases.

# ABCB1 is known for its role in drug transport and multidrug resistance, meaning it might be linked to diverse conditions rather than a specific disease class.

# to show evidence type, filter out clinical or variant, focus with cancer
```

6. Analyze cancer
```{r}
# Filter cancer-related diseases and exclude "clinical" or "variant" evidence types
cancer_evidence <- filtered_relations %>%
  mutate(Category = case_when(
    grepl("cancer|carcinoma|lymphoma|tumor", entity2_name, ignore.case = TRUE) ~ "Cancer",
    grepl("diabetes", entity2_name, ignore.case = TRUE) ~ "Metabolic",
    grepl("hypertension|cardiac|coronary|arrhythmia", entity2_name, ignore.case = TRUE) ~ "Cardiovascular",
    grepl("autism|attention deficit|nervous system", entity2_name, ignore.case = TRUE) ~ "Neurological",
    TRUE ~ "Other"
  )) %>%
  filter(Category == "Cancer")
```


7. Get the Table for cancer associated with its evidence
```{r}
cancer_disease_evidence <- cancer_evidence %>%
  select(entity2_name, evidence) %>%
  arrange(entity2_name)

# View the first few rows
head(cancer_disease_evidence)

# Count number of times each (disease, evidence) pair appears
cancer_disease_evidence_counts <- cancer_evidence %>%
  count(entity2_name, evidence, sort = TRUE)

# Check the counts
head(cancer_disease_evidence_counts)
```

8. Plotting the result
```{r}
ggplot(cancer_disease_evidence_counts, aes(x = reorder(entity2_name, n), y = n, fill = evidence)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Cancer-Related Diseases Vs. Evidence Types",
       x = "Disease",
       y = "Number of Associations",
       fill = "Evidence Type") +
  scale_fill_brewer(palette = "Paired")+
  theme_light()

# Basically different types of Lymphoma are clinical+variant associated, and carcinoma is variant associated.
```
```{r}
ggplot(cancer_disease_evidence_counts, aes(x = evidence, y = n, fill = evidence)) +
  geom_bar(stat = "identity") +
  labs(title = "Cancer-Related Diseases Vs. Evidence Types",
       x = "Disease",
       y = "Number of Associations",
       fill = "Evidence Type") +
  scale_fill_brewer(palette = "Paired")+
  theme_light()

#Gives alternate more of quantitative result: how many of these associations are each type of evidence
```
