---
title: "PharmGKB Analysis"
output: html_document
date: "2025-03-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Goals
1. How does Gene ABCB1 associate with different diseases?
2. How strong is the evidence for the association between Variant X and different disease?  

##Resources  
- [Warfarin data](https://www.pharmgkb.org/downloads)  

## Load the libraries
```{r}
library(tidyverse)
library(janitor)
library(dplyr)
```

#Data
```{r}
relations <- read_tsv("data/relationships.tsv") %>% clean_names()
```

#Dataframe with only variant genes and diseases
```{r}
variant_disease <- relations %>% 
  filter(entity1_type == "Variant" & entity2_type == "Disease")
```

#Percentage of association between variant genes and diseases
```{r}
variant_disease %>% 
  group_by(association) %>%
  summarise(count = n(), .groups = 'drop') %>%
  mutate(percentage = (count / sum(count)) * 100) %>% 
  arrange(-count)
```

#Percentage of association between regular genes and diseases
```{r}
relations %>% 
  filter(entity1_type == "Gene" & entity2_type == "Disease") %>% 
  group_by(association) %>%
  summarise(count = n(), .groups = 'drop') %>%
  mutate(percentage = (count / sum(count)) * 100) %>% 
  arrange(-count)
```

#Which gene is most associated with having a relationship with a disease?
```{r}
variant_disease %>% 
  group_by(entity1_name) %>%
  filter(association == "associated") %>% 
  summarise(count = n(), .groups = 'drop') %>%
  mutate(percentage = (count / sum(count)) * 100) %>% 
  arrange(-percentage)
```
The presence of the variant rs1045642 gene is most associated with having a disease out of all the other variant genes.

#Graph of genes most associated with having a relationship with a disease
```{r}
variant_disease %>% 
  group_by(entity1_name, association) %>%
  summarise(count = n(), .groups = 'drop') %>%
  mutate(percentage = (count/sum(count)) * 100) %>%
  arrange(-count) %>%
  slice_max(order_by = count, n = 10) %>%
  ggplot(aes(x = reorder(entity1_name, percentage), y = percentage, fill = association)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  labs(title = "Association Status by Variant Gene",
       x = "Variant Gene",
       y = "Percentage (%)",
       fill = "Association Type") +
  scale_fill_brewer(palette = "Paired")
  theme_light()
```
#Which diseases is the rs1045642 gene associated with?
```{r}
variant_disease %>% 
  select(entity1_name, entity2_name, association) %>% 
  filter(entity1_name == "rs1045642" & association == "associated")
```
#What diseases are most associated with variant genes?
```{r}
variant_disease %>% 
  group_by(entity2_name) %>%
  filter(association == "associated") %>% 
  summarise(count = n(), .groups = 'drop') %>%
  mutate(percentage = (count / sum(count)) * 100) %>% 
  arrange(-percentage)
```

#Graph of most associated diseases 
```{r}
variant_disease %>% 
  group_by(entity2_name, association) %>%
  summarise(count = n(), .groups = 'drop') %>%
  mutate(percentage = (count/sum(count)) * 100) %>%
  arrange(-count) %>%
  slice_max(order_by = count, n = 10) %>%
  ggplot(aes(x = reorder(entity2_name, percentage), y = percentage, fill = association)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  labs(title = "Association Status by Variant Gene",
       x = "Disease",
       y = "Percentage (%)",
       fill = "Association Type") +
  scale_fill_brewer(palette = "Paired")
  theme_light()
```

#Comparing amount of evidence for associated vs. not associated interactions
```{r}
ev_assoc <- variant_disease %>% 
  separate_rows(pmi_ds, sep = ";") %>% 
  group_by(association) %>%
  summarize(pmid_count = n())

head(ev_assoc)
```


