---
title: "PharmGKB Analysis"
output: html_document
date: "2025-03-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Goals
1. How does Gene ABCB1 associate with different diseases?
2. How strong is the evidence for the association between Variant X and different disease?  

##Resources  
- [Warfarin data](https://www.pharmgkb.org/downloads)  

## Load the libraries
```{r}
library(tidyverse)
library(janitor)
library(dplyr)
```

#Data
```{r}
relations <- read_tsv("data/relationships.tsv") %>% clean_names()
```

#Dataframe with only variant genes and diseases
```{r}
variant_disease <- relations %>% 
  filter(entity1_type == "Variant" & entity2_type == "Disease")
```

#Percentage of association between variant genes and diseases
```{r}
pa_variant_disease <- variant_disease %>% 
  group_by(association) %>%
  summarise(count = n(), .groups = 'drop') %>%
  mutate(percentage = (count / sum(count)) * 100) %>% 
  arrange(-count)
pa_variant_disease 
```

#Graph of percentage of association between variant genes and diseases
```{r}
ggplot(pa_variant_disease, aes(x = "", y = percentage, fill = association)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  labs(title = "Percentage of Association Types between Variant Genes and Diseases",
       fill= "Type of Association") +
  scale_fill_brewer(palette = "Paired")+
  theme_void()
```


#Which gene is most associated with having a relationship with a disease?
```{r}
variant_disease %>% 
  group_by(entity1_name) %>%
  filter(association == "associated") %>% 
  summarise(count = n(), .groups = 'drop') %>%
  mutate(percentage = (count / sum(count)) * 100) %>% 
  arrange(-percentage)
```
The presence of the variant rs1045642 gene is most associated with having a disease out of all the other variant genes.

#Graph of genes most associated with having a relationship with a disease
```{r}
variant_disease %>% 
  group_by(entity1_name, association) %>%
  summarise(count = n(), .groups = 'drop') %>%
  mutate(percentage = (count/sum(count)) * 100) %>%
  arrange(-count) %>%
  slice_max(order_by = count, n = 10) %>%
  ggplot(aes(x = reorder(entity1_name, percentage), y = percentage, fill = association)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  labs(title = "Association Status by Variant Gene",
       x = "Variant Gene",
       y = "Percentage (%)",
       fill = "Association Type") +
  scale_fill_brewer(palette = "Paired")+
  theme_light()
```

#Which diseases is the rs1045642 gene associated with?
```{r}
variant_disease %>% 
  select(entity1_name, entity2_name, association) %>% 
  filter(entity1_name == "rs1045642" & association == "associated")
```

#What diseases are most associated with variant genes?
```{r}
variant_disease %>% 
  group_by(entity2_name) %>%
  filter(association == "associated") %>% 
  summarise(count = n(), .groups = 'drop') %>%
  mutate(percentage = (count / sum(count)) * 100) %>% 
  arrange(-percentage)
```

#Graph of most associated diseases 
```{r}
variant_disease %>% 
  group_by(entity2_name, association) %>%
  summarise(count = n(), .groups = 'drop') %>%
  mutate(percentage = (count/sum(count)) * 100) %>%
  arrange(-count) %>%
  slice_max(order_by = count, n = 10) %>%
  ggplot(aes(x = reorder(entity2_name, percentage), y = percentage, fill = association)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  labs(title = "Association Status by Variant Gene",
       x = "Disease",
       y = "Percentage (%)",
       fill = "Association Type") +
  scale_fill_brewer(palette = "Paired")+
  theme_light()
```

#Categorize diseases: MODIFY- innacurate
```{r}
disease_count_v <- variant_disease %>%
  mutate(Category = case_when(
    grepl("cancer|carcinoma|lymphoma|tumor", entity2_name, ignore.case = TRUE) ~ "Cancer",
    grepl("diabetes", entity2_name, ignore.case = TRUE) ~ "Metabolic",
    grepl("hypertension|cardiac|coronary|arrhythmia", entity2_name, ignore.case = TRUE) ~ "Cardiovascular",
    grepl("autism|attention deficit|nervous system", entity2_name, ignore.case = TRUE) ~ "Neurological",
    TRUE ~ "Other"
  ))

category_counts <- disease_count_v %>%
  count(Category, sort = TRUE)

ggplot(category_counts, aes(x = reorder(Category, n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Disease Categories Associated with Variant Genes",
       x = "Disease Category",
       y = "Number of Associations") +
  theme_light()
```

#Comparing amount of evidence for associated vs. not associated interactions
```{r}
ev_assoc <- variant_disease %>% 
  separate_rows(pmi_ds, sep = ";") %>% 
  group_by(association) %>%
  summarize(pmid_count = n())

head(ev_assoc)
```

#Analysis of clinical vs. variant annotation
```{r}
annotation_count_vd <- variant_disease %>% 
  filter(association == "associated") %>% 
  group_by(evidence) %>%
  summarise(count = n(), .groups = 'drop') %>%
  mutate(percentage_e = (count / sum(count)) * 100) %>% 
  arrange(-count)
annotation_count_vd
```
#Cancer

7. Get the Table for cancer associated with its evidence
```{r}
# Filter cancer-related diseases and exclude "clinical" or "variant" evidence types
cancer_evidence <- variant_disease %>%
  mutate(Category = case_when(
    grepl("cancer|carcinoma|lymphoma|tumor", entity2_name, ignore.case = TRUE) ~ "Cancer",
    grepl("diabetes", entity2_name, ignore.case = TRUE) ~ "Metabolic",
    grepl("hypertension|cardiac|coronary|arrhythmia", entity2_name, ignore.case = TRUE) ~ "Cardiovascular",
    grepl("autism|attention deficit|nervous system", entity2_name, ignore.case = TRUE) ~ "Neurological",
    TRUE ~ "Other"
  )) %>%
  filter(Category == "Cancer")
```

```{r}
cancer_disease_evidence <- cancer_evidence %>%
  select(entity2_name, evidence) %>%
  arrange(entity2_name)

# View the first few rows
head(cancer_disease_evidence)

# Count number of times each (disease, evidence) pair appears
cancer_disease_evidence_counts <- cancer_evidence %>%
  count(entity2_name, evidence, sort = TRUE)

# Check the counts
head(cancer_disease_evidence_counts)
```
```{r}
ggplot(cancer_disease_evidence_counts, aes(x = reorder(entity2_name, n), y = n, fill = evidence)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Cancer-Related Diseases Vs. Evidence Types",
       x = "Disease",
       y = "Number of Associations",
       fill = "Evidence Type") +
  scale_fill_brewer(palette = "Paired")+
  theme_light()

# Basically different types of Lymphoma are clinical+variant associated, and carcinoma is variant associated.
```

```{r}
ggplot(annotation_count_vd, aes(x = evidence, y = count, fill = evidence)) +
  geom_bar(stat="identity") +
  labs(title = "Cancer-Related Diseases Vs. Evidence Types",
       x = "Disease",
       y = "Number of Associations",
       fill = "Evidence Type") +
    theme(axis.text.x = element_text(angle = 60, hjust = 1))+
    scale_fill_brewer(palette = "Paired")

#Gives alternate more of quantitative result: how many of these associations are each type of evidence
```

#assign priority/ranking to each evidence type based off of clinical vs. variant vs. both annotation — make a plot based off of that (NOT BAR)


